### 2015-07-21 implementation-concretization.tex
Tighten text.

Spack's concretization algorithm is shown in Figure~\ref{fig:concretization}.
The process starts when a user invokes {\tt spack install} and requests that a
spec be built.  Spack converts the spec to an abstract DAG.
It then builds a {\it separate} spec DAG for any constraints that are encoded
in directives in package files.

Spack intersects the constraints of the two DAGs package by package, and it
checks each parameter for inconsistencies.  Inconsistencies can arise if, for example,
the user inadvertently requests two versions of the same package, or if a
package file explicitly requests a different version than the user requested.
Likewise, if the package and the user specified different compilers, variants,
or platforms for particular packages, this will cause the user to be notified
of the conflict. If the user or package specified version ranges, they are
intersected, and if the ranges do not overlap, an error will be raised.
When the intersection succeeds, Spack has a single spec DAG with the merged
constraints of the user and the package files.

The next part of the process is iterative.
If any node in the DAG is a virtual dependency, spack replaces it with with a
suitable interface provider.  It does this by building a reverse
index from virtual packages to providers using the {\tt provides when}
directives discussed in Section~\ref{sec:virtual}. If multiple providers
could satisfy the constraints on the virtual spec,
Spack will consult site and user policies to select the ``best'' possible
provider.  The selected provider may {\it itself} have virtual dependencies,
so Spack repeats this process until there are no more virtual packages
in the DAG.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
	\caption{
		Concretized spec from Figure~\ref{fig:specs-mpileaks}.
		\label{fig:specs-mpileaks-concrete}
	}
\end{figure}
When the concretization process completes, it outputs a fully concrete spec DAG.
Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures,
compilers, versions, variants, and all dependencies resolved.
This fulfills Spack's guarantees.

At install time, Spack constructs a package object for each node in the spec DAG
and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
{\tt install} method.  For the {\tt spec} parameter to {\tt install}, it passes
a sub-DAG rooted at current node, which is also a concrete spec.  Package authors are
responsible for querying the spec in {\tt install} and handling different configurations.

@@

Figure~\ref{fig:concretization-time} shows the running time of concretization vs.
the size of a package DAG (in nodes).  We generated this plot by running {\tt concretize}
on all 245 packages in our internal version of Spack.  We see that for all but the 9 largest
packages, the algorithm takes less than 2 seconds on a Sandy Bridge cluster node, and less
than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin to see superlinear
scaling, but even for 50 nodes or more, concretization takes less than 6 seconds.
%
This performance is expected from a greedy, fixed-point method.
The algorithm repeats, evaluating conditions and adding dependencies and constraints, until
there are no more changes to be made.
%
%  Additionally, our current compiler concretization
% policy requires a search in the DAG; it prefers the compiler of the nearest ancestor
% if the current node's compiler is not set. 
%
More generally, concretization is an instance of the constraint
satisfaction problem, which is NP-complete. Our greedy solver is sufficient for the
packages we have built so far, and its running time is currently negligible. We do
not expect to see software packages with thousands of dependencies in the near
future, and we do not expect concretization to become a bottleneck, even if we
use a full constraint solver.

>>>>

Spack's concretization algorithm is shown in Figure~\ref{fig:concretization}.
The process starts when a user invokes {\tt spack install} and requests that a
spec be built.  Spack converts the spec to an abstract DAG.
It then builds a {\it separate} spec DAG for any constraints encoded
by directives in package files.
%
Spack intersects the constraints of the two DAGs package by package, and it
checks each parameter for inconsistencies.  Inconsistencies can arise if, for example,
the user inadvertently requests two versions of the same package, or if a
package file depends on a different version than the user requested.
Likewise, if the package and the user specified different compilers, variants,
or platforms for particular packages, Spack will stop and notify
the user of the conflict. If the user or package specifies version ranges,
they are intersected, and if the ranges do not overlap, an error is raised.
When the intersection succeeds, Spack has a single DAG with the merged
constraints of the user and the package files.

The next part of the process is iterative.
If any node is a virtual dependency, spack replaces it with with a
suitable interface provider.  It does this by building a reverse
index from virtual packages to providers using the {\tt provides when}
directives (Section~\ref{sec:virtual}). If multiple providers
satisfy the virtual spec's constraints,
Spack consults site and user policies to select the ``best'' possible
provider.  The selected provider may {\it itself} have virtual dependencies,
this process is repeated until there are no more virtual packages in the DAG.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
	\caption{
		Concretized spec from Figure~\ref{fig:specs-mpileaks}.
		\label{fig:specs-mpileaks-concrete}
	}
\end{figure}
On completion, the concretization outputs a fully concrete spec DAG.
Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures,
compilers, versions, variants, and all dependencies resolved.
This fulfills Spack's guarantees.
%
At install time, Spack constructs a package object for each node in the spec DAG
and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
{\tt install} method.  For the {\tt spec} parameter to {\tt install}, it passes
a sub-DAG rooted at current node (also a concrete spec).  Package authors must
query the spec in {\tt install} to handle different configurations.

@@

Figure~\ref{fig:concretization-time} shows the running time of concretization vs.
package DAG size in nodes.  We generated this plot by running {\tt concretize}
on all of Spack's 245 packages.  We see that for all but the 9 largest
packages, concretization takes less than 2 seconds on a Sandy Bridge machine, 
and less than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin
to see superlinear scaling, but even for 50 nodes or more, concretization
takes less than 6 seconds.
%
This performance is expected from a greedy, fixed-point method.
The algorithm repeats, evaluating conditions and adding dependencies and constraints, until
there are no more changes to be made.
%
%  Additionally, our current compiler concretization
% policy requires a search in the DAG; it prefers the compiler of the nearest ancestor
% if the current node's compiler is not set. 
%
More generally, concretization is an instance of the constraint
satisfaction problem, which is NP-complete. Our greedy solver is sufficient for the
packages we have built so far, and its running time is currently negligible. We do
not expect to see software packages with thousands of dependencies in the near
future, and we do not expect concretization to become a bottleneck, even if we
use a full constraint solver.

### 2015-07-21 implementation-specs.tex
Install environment revisions.
Name change spack-paper-sc15 -> spack-sc15
Tighten text.

Using the simple script in Figure~\ref{fig:mpileaks}, Spack can build many different
versions and configurations of the {\tt mpileaks} package.  In traditional port systems,
package code is structured to build a single version of a package, but in Spack, each
package file is a {\it template} that can be configured and built in many
different ways, according to a set of {\it parameters}.
Spack calls a single build configuration a {\it spec}. The {\tt spec} object
passed to {\tt install()} is how Spack communicates dependencies and parameters to package authors.

\subsubsection{Structure}
To understand how specs work, consider the {\tt mpileaks} package structure.
Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
its relationships with other packages.  There are two direct dependencies:
the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
for each dependency and constructs a graph of their relationships.  The result
is a directed, acyclic graph (DAG) of dependencies.\footnote{Spack currently disallows
circular dependencies.}
%
To guarantee a consistent build, and to avoid ABI compatibility issues, the DAG
is constructed so that there is only {\it one} version of any particular package.  Note
the distinction: while Spack can install arbitrarily many configurations of any package,
no two configurations of the same package will ever appear in the same build DAG.

@@

We begin with a simple example.
Consider the case where a user wants to install the {\tt mpileaks} package, but knows
nothing about its structure.  To install the package, the user invokes {\tt spack install}:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks
\end{minted}
%
Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
Note that even though the spec contains no dependency information, it is still
converted to a full DAG, based on directives supplied in package files. Since there
are no constraints on the nodes, Spack has considerable leeway when building the
package, and we say that the package is {\it unconstrained}.
%
Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
with a version constraint after the package name:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3
\end{minted}
%
We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
If the user does not need a specific version of {\tt mpileaks}, but does require
particular minimum version, then the user could use {\it version range} syntax
and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save time if
Spack already has a version installed that satisfies the spec -- Spack will use
the previously-built installation instead of building a new one.

Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of the spec syntax:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
\end{minted}
%
The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
we see that there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
and that the user has requested the {\tt debug} variant of the {\tt callpath} library.

@@

Table~\ref{tab:specs} shows further examples of specs, ranging from very
simple to more complex. From these examples, we can see that Spack offers constraint
notation to cover the rest of the HPC package parameter space.

{\bf Versions.}
Version constraints, denoted with {\tt @}, were already covered above. Versions
can be precise ({\tt @2.5.1}) or denote a range ({\tt @2.5:4.4}),
which may be open-ended ({\tt @2.5:}).

The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
in our experience users frequently want bleeding-edge versions, and package managers
frequently lag behind the latest releases.
Spack has a capability to extrapolate URLs from versions,
using the package's {\tt url} attribute as a model.\footnote{This works
for packages with consistently named URLs.}  The user can request a specific
version on the command line, even if it is unknown to Spack,
and Spack will attempt to install it.  Spack also uses the same
model to scrape webpages and find new versions as they become available.

{\bf Compilers.}
With a compiler constraint (shown on line 3) the user
simply adds {\tt \%} followed by its name, along with an optional compiler version
specifier.  Spack compiler names, e.g. {\tt gcc}, refer to the full compiler {\it toolchain},
i.e. the C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
compiler toolchains if they are in the user's {\tt PATH}, or they can be registered manually
through a configuration file.

{\bf Variants.}
To handle build options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must check the spec and handle the cases
where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
or {\tt \textasciitilde \ignorespaces debug}).  The names simplify the versioning and prevent
Spack's configuration space from becoming too fine-grained.
It would violate our goal of conciseness, for example, to include detailed
compiler flags in spec syntax, but known sets of flags can simply be named.

{\bf Cross-compilation.}
To support cross-compilation, Spack includes the platform in the package spec (line 5).
Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
specified per-package; this allows front-end tools to depend on their back-end measurement
libraries with a {\it different} architecture on cross-compiled machines.

@@

So far, we have shown examples of specs being used to request constraints from the
command line, when {\tt spack install} is invoked.  However, the user is not the only
source of constraints.  Applications may require specific versions of dependencies,
and it is often desirable to write constraints like this into a package file.  For
example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
At first glance, the {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} look
like they take simple package names.  However, a package name is also a spec, and
the same constraint syntax usable from the command line can be applied inside directives.
So, the ROSE compiler can simply write:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.54.0')
\end{minted}
%
This constraint will be incorporated into the initial DAG node generated from
the {\tt ROSE} package.  Dependencies can also be {\it conditional}.  For example,
the {\tt ROSE} package builds with different versions of boost, depending on the
compiler version. To handle this, directives also accept spec syntax as a
predicate in an optional {\tt when} parameter:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.47.0', when='%gcc@:4')
    depends_on('boost@1.54.0', when='%gcc@5:')
\end{minted}
%
The same notation is used to optionally build with libraries like MPI:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('mpi', when='+mpi')
\end{minted}
%
and to ensure that specific patches are applied to Python source code
when it is built on Blue Gene/Q, with different compilers:
%
\begin{minted}[fontsize=\scriptsize]{python}
    patch('python-bgq-xlc.patch',   when='=bgq%xl')
    patch('python-bgq-clang.patch', when='=bgq%clang')
\end{minted}
%
Constraints in the {\tt when} clause are matched against the package spec.
Outside of directives, constraints can also be tested directly on the {\tt spec}
object in the {\tt install} method:
%
\begin{minted}[fontsize=\scriptsize]{python}
  def install(self, spec, prefix):
      if spec.satisfies('%gcc'):
          # Handle gcc
      elif spec.satisfies('%xl'):
          # Handle XL compilers
      ...
\end{minted}
%

@@

%class Dyninst(Package):
%    """"Dyninst binary instrumentation and analysis tool."""
%    url="http://www.paradyn.org/release8.1/DyninstAPI-8.1.1.tgz"
%    version('8.1.1', 'd1a04e995b7aa70960cd1d1fac8bd6ac')
%    depends_on("libelf")
%    depends_on("libdwarf")
%    depends_on("boost@1.42:")
%
\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
    def install(self, spec, prefix):  # default build uses cmake
        with working_dir('spack-build', create=True):
            cmake('..', *std_cmake_args)
            make()
            make("install")

>>>>

Using the simple script in Figure~\ref{fig:mpileaks}, Spack can build many different
versions and configurations of the {\tt mpileaks} package.  In traditional port systems,
package code is structured to build a single version of a package, but in Spack, each
package file is a {\it template} that can be configured and built in many
different ways, according to a set of {\it parameters}.
Spack calls a single build configuration a {\it spec}. The {\tt spec} object
passed to the {\tt install} method is how Spack communicates dependencies and parameters to package authors.

\subsubsection{Structure}
To understand specs, consider the {\tt mpileaks} package structure.
Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
its relationships with other packages.  There are two direct dependencies:
the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
for each dependency and constructs a graph of their relationships.  The result
is a directed, acyclic graph (DAG).\footnote{Spack currently disallows
circular dependencies.}
%
To guarantee a consistent build and to avoid ABI compatibility issues, we construct
the DAG so that there is only {\it one} version of any package.  Note
the distinction: while Spack can install arbitrarily many configurations of any package,
no two configurations of the same package will ever appear in the same build DAG.

@@

We begin with a simple example.
Consider the case where a user wants to install the {\tt mpileaks} package, but knows
nothing about its structure.  To install the package, the user invokes {\tt spack install}:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks
\end{minted}
%
Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
Spack converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
Note that even though the spec contains no dependency information, it is still
converted to a full DAG, based on directives supplied in package files. Since there
are no constraints on the nodes, Spack has leeway when building the
package, and we say that it is {\it unconstrained}.
%
Suppose instead that the user wants a specific {\tt mpileaks} version.
She can request it with a version constraint after the package name:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3
\end{minted}
%
We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
If the user does not need a specific version of {\tt mpileaks}, but does require
particular minimum version, then the user could use {\it version range} syntax
and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save time if
Spack already has a version installed that satisfies the spec -- Spack will use
the previously-built installation instead of building a new one.

Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of the spec syntax:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
\end{minted}
%
The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
and the user has requested the {\tt debug} variant of the {\tt callpath} library.

@@

Table~\ref{tab:specs} shows further examples of specs, ranging from very
simple to more complex. These examples show how Spack's constraint
notation covers the rest of the HPC package parameter space.

{\bf Versions.}
Version constraints are denoted with {\tt @}. Versions
can be precise ({\tt @2.5.1}) or denote a range ({\tt @2.5:4.4}),
which may be open-ended ({\tt @2.5:}).
%
The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with
checksums, but
in our experience users frequently want bleeding-edge versions, and package managers
often lag behind the latest releases.
Spack can extrapolate URLs from versions,
using the package's {\tt url} attribute as a model.\footnote{This works
for packages with consistently named URLs.}  If the user requests a specific
version on the command line that is unknown to Spack,
Spack will attempt to fetch and install it.  Spack also uses the same
model to scrape webpages and find new versions as they become available.

\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
    def install(self, spec, prefix):  # default build uses cmake
        with working_dir('spack-build', create=True):
            cmake('..', *std_cmake_args)
            make()
            make("install")

{\bf Compilers.}
With a compiler constraint (shown on line 3) the user
simply adds {\tt \%} followed by its name, with an optional compiler version
specifier.  Spack compiler names like {\tt gcc} refer to the full compiler {\it toolchain},
i.e. the C, C++, and Fortran 77 and 90 compilers.  Spack can auto-detect
compiler toolchains in the user's {\tt PATH}, or they can be registered manually
through a configuration file.

{\bf Variants.}
To handle options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must explicitly handle cases
where debug is enabled ({\tt +debug}) or disabled ({\tt -debug}
or {\tt \textasciitilde \ignorespaces debug}).  Names simplify versioning
and prevent the configuration space from becoming too large.
For example, including detailed compiler flags in spec syntax
would violate our goal of conciseness, but known sets of flags can simply be named.

{\bf Cross-compilation.}
To support cross-compilation, specs includes the system architecture (line 5).
Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
specified per-package; this allows front-end tools to depend on their back-end measurement
libraries with a {\it different} architecture on cross-compiled machines.

@@

So far, we have shown examples of specs used to request constraints from the
command line, when {\tt spack install} is invoked.  However, the user is not the only
source of constraints.  Applications may require specific versions of dependencies,
and it is often desirable to write constraints like this into a package file.  For
example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
The {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} contain
simple package names.  However, a package name is also a spec, and
the same constraint syntax usable from the command line can be applied inside directives.
So, the ROSE compiler can simply write:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.54.0')
\end{minted}
%
This constraint will be incorporated into the initial DAG node generated from
the {\tt ROSE} package.  Dependencies can also be {\it conditional}.  For example,
the {\tt ROSE} package builds with different versions of boost, depending on the
compiler version. To handle this, directives also accept spec syntax as a
predicate in an optional {\tt when} parameter:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.47.0', when='%gcc@:4')
    depends_on('boost@1.54.0', when='%gcc@5:')
\end{minted}
%
The same notation is used to optionally build with libraries like MPI:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('mpi', when='+mpi')
\end{minted}
%
and to ensure that specific patches are applied to Python source code
when it is built on Blue Gene/Q, with different compilers:
%
\begin{minted}[fontsize=\scriptsize]{python}
    patch('python-bgq-xlc.patch',   when='=bgq%xl')
    patch('python-bgq-clang.patch', when='=bgq%clang')
\end{minted}
%
Constraints in the {\tt when} clause are matched against the package spec.
Outside of directives, constraints can also be tested directly on the {\tt spec}
object in the {\tt install} method:
%
\begin{minted}[fontsize=\scriptsize]{python}
  def install(self, spec, prefix):
      if spec.satisfies('%gcc'):
          # Handle gcc
      elif spec.satisfies('%xl'):
          # Handle XL compilers
      ...
\end{minted}
%

@@

%class Dyninst(Package):
%    """"Dyninst binary instrumentation and analysis tool."""
%    url="http://www.paradyn.org/release8.1/DyninstAPI-8.1.1.tgz"
%    version('8.1.1', 'd1a04e995b7aa70960cd1d1fac8bd6ac')
%    depends_on("libelf")
%    depends_on("libdwarf")
%    depends_on("boost@1.42:")
%
